---
title: "A grammar of data wrangling: Part 2 "
subtitle: "STAT 311"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output: html_document
---


We now return to the `gapminder` data frame to answer more questions using the data verbs we learned today: count`, group_by` and `summarise`. 

Let's load the packages we will need today by running the following code chunk using the green Play button. 

```{r setup, include = FALSE}
#load packages and set global options
library(tidyverse)        #umbrella package
library(gapminder)        #dataset is in this package

```

## Warm Up

Load the data `gapminder` by typing `data(gapminder)` function in the Console and take a look at it in the spreadsheet editor.


* * * 


## Part 2: count, group_by, summarise

Fill in the blanks in the code provided in each code chunk to accomplish a specific task. Consult the slide deck "Introduction to Data: Data Manipulation" for help. Each code chunk has the `eval` option set to FALSE. Once you have filled in the blanks you may delete this option so your code will run when the document is knitted.

### Exercises


### Exercise 1

In the following code chunk, I have created a new dataframe called `gapminder_countries` which lists the distinct countries in `gapminder` and the continent they are located in.

Run this code chunk and look at the resulting dataframe. 

```{r country}

gapminder_countries <- gapminder %>% 
  select(country, continent) %>%
  distinct(country, .keep_all = TRUE)
                          
```

Which continent has the most countries in it? Which has the least?

```{r ex1, eval = FALSE}

gapminder_countries %>% 
  count(___)

```

The remaining exercises refer to the original `gapminder` data frame.

### Exercise 2

Summarize the life expectancy `lifeExp`  by calculating the mean.

```{r ex2, eval = FALSE}

gapminder %>% 
  summarise( mean_lifeExp = ___ )

```


### Exercise 3

Summarize life expectancy `lifeExp` in Africa in 1952 by calculating the mean.

```{r ex3, eval = FALSE}

gapminder %>% 
  filter( ___) %>%
  summarise( mean_lifeExp = ___ )

```


### Exercise 4

Summarize life expectancy `lifeExp` in 1952 for each continent separately by calculating the mean.

```{r ex4, eval = FALSE}

gapminder %>%
  filter( year [EQUAL TO] ___) %>%
  group_by(___) %>%
  summarise( ___ )


```



### Exercise 5

Summarize life expectancy `lifeExp` in 1952 for each continent separately by calculating the mean and median and sort the results in descending order by median life expectancy. 


```{r ex5, eval = FALSE}

gapminder %>% 
  filter(___) %>%
  group_by( ___) %>%
  summarise( mean_lifeExp = ___, 
             median_lifeExp = ___ ) %>%
  arrange( ___ )

```


### Exercise 6

Create a new variable called `popBil` which is the population in billions. For example, 8425333 is 842533/1000000000 = .0008425333 billions.  Then use this new variable to calculate the total population in the world for each year. 

```{r ex6, eval = FALSE}

gapminder %>% 
      mutate( popBil = ___ ) %>% 
      group_by(___) %>% 
      summarise( total_pop = sum(___) ) 

```


**Participation 2: Please do your calculations for the remaining problems below and then record your answers in CANVAS for the quiz titled Participation 2**



### Exericise 7

Summarise the life expectancy `lifeExp` in 1952 by calculating the mean.


### Exercise 8

Summarise the life expectancy `lifeExp` in Asia in 1952 by calculating the mean. 

### Exercise 9

Summarise the life expectancy `lifeExp` for each continent in 2002 by calculating the median. Sort your results in descending order.


### Exercise 10

Create a new variable called `gdp` which recovers each country's GDP by multiplying the population `pop` and GDP per capita `gdpPercap`. Summarise the `gdp` for each country by calculating the total across all the years. Which 3 countries have the highest totals?

 

